---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r message=FALSE }
# Load libraries
library(dplyr)
library(lubridate)

```
<h6> Load helper functions (they are printed at the end of the document) </h6>
* convert_interval_values [0 5 10 ... 55 100 105 ... converts to 0 5 10 ... 55 60 65 ... ]
* convert_interval_back2original [55 60 65 ... converts to 55 100 105 ... ]
* imputedata [a suitable dataframe that works as a lookuptable and a dataframe returns imputed dataframe ]
```{r echo=FALSE}
# Helper-functions OBS! will be echoed in the end of the document
# we have gap in the timeseries. Motivation see playground.R
convert_interval_values <- function(interval){
  # 0 5 .... 55 100 105 ... converts to 0 5 .... 55 60 65 ...
  a <- interval %/% 100
  b <- interval %% 100 
  res <- a * 60 + b
  return(res)
}

# Convert back to original value on interval
convert_interval_back2original <- function(interval) {
  # 55 converts to 55
  # 60 converts to 100
  # 125 converts to 205
  a <- interval %/% 60
  b <- interval %% 60 
  res <- a * 100 + b
  return(res)
  
}

# # lut provides key - value pairs to use to impute "steps" data
# imputedata <- function(lut , df) {
#   # lut is a dataframe providing suitable replacement for missing values 
#   names(lut) <- c("key", "value")
#   na_ind <- is.na(df$steps)
#   intervals2impute <- df[na_ind, "interval"]
#   to_impute <- lut[intervals2impute %in% lut$key ,"value"]
#   df$steps[na_ind] <- to_impute
#   return(df)
#   #to_impute
# }

imputedata <- function(lut , df) {
  # lut is a dataframe providing suitable replacement for missing values 
  names(lut) <- c("key", "value")
  na_ind <- is.na(df$steps)
  intervals2impute <- df[na_ind, "interval"]
  
  myvar_vec <- vector()
  for(i in seq_along(intervals2impute[[1]])) {
    e <- intervals2impute[[1]][i]
    myvar_vec[i] <- lut[lut$key==e,]$value
  }
  df_temp <- df
  df_temp[na_ind,]$steps <- myvar_vec[[1]]
  df_temp
}


```
```{r echo=FALSE}
# # Load data (if on local computer)
# if(!file.exists("activity.zip")) {
#   url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
#   dest <- "activity.zip"
#   download.file(url = url, destfile = dest)
#   # Uncomment if on local computer 
#   #unzip(zipfile = "activity.zip")
# }
# df <- read.table(file = "activity.csv", header = TRUE, 
#                  colClasses = colClasses, sep="," )


```
<h3> Load data </h3>
```{r}
# If on github (else if on local computer, see code commented out or in playground.r)
if(!file.exists("activity.zip")) {
  print("activity.zip not found")
  print("if on local computer uncomment code above and comment out this codechunk")
  stop("missing activity.zip in working directory")
}
colClasses <- c("numeric", "Date", "numeric")
# For use on github (don't unzip)
df <- read.table(file = unz("activity.zip", "activity.csv"), header = TRUE, 
                 colClasses = colClasses, sep="," )


```
<h3> Transform data to suitable format (tbl_df and changes in interval) </h3>
```{r}

df <- tbl_df(df)
df$interval2 <- convert_interval_values(df$interval)


# Let's remove the original interval
df <- df %>%
  select(steps, date, interval = interval2)


```

## What is mean total number of steps taken per day?

```{r}
# Calculate the total number of steps taken per day
df_tot <- df %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps, na.rm = TRUE))
```

```{r}
# histogram of the total number of steps taken each day
hist(df_tot$total_steps, breaks = 20, xlab = "Steps per day")
```

```{r}
#mean and median of the total number of steps taken per day
mean_and_median <- df_tot %>%
  summarize(avg_steps_per_day = mean(total_steps, na.rm = TRUE), 
          median_steps_per_day = median(total_steps, na.rm = TRUE))

print(paste("Mean steps/day: ",round(mean_and_median$avg_steps_per_day,digits = 1)))
print(paste("Median steps/day: ",mean_and_median$median_steps_per_day))


# mean and median steps/day
# OBS! dates with only NA in steps will have result 0 in the summarize
# another option would be to remove steps with NA as value --> less dates

```


## What is the average daily activity pattern?

```{r}
df_interval <- df %>%
  group_by(interval) %>%
  summarize(avg_steps_per_interval = mean(steps, na.rm = TRUE))




ind <- which(df_interval$avg_steps_per_interval == max(df_interval$avg_steps_per_interval))
max_int <- df_interval[ind, "interval"][[1]]

with(df_interval,plot(x= interval, 
                      y = avg_steps_per_interval, type="l"))
abline(v = max_int, lwd = 2, col = "red")

```

```{r}
# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
str1 <- paste("Max steps on average was taken in interval = ", max_int, " indicated by red line ")
str2 <- paste("or in original interval unit : ", convert_interval_back2original(max_int) )
print(str1)
print(str2)


```
## Imputing missing values
```{r}
df_interval_median <- df %>%
  group_by(interval) %>%
  summarize(median_steps_per_interval = median(steps, na.rm = TRUE))

df_interval_max <- df %>%
  group_by(interval) %>%
  summarize(max_steps_per_interval = max(steps, na.rm = TRUE))

df_imputed1 <- imputedata(lut = df_interval_median, df)

df_imputed2 <- imputedata(lut = df_interval_max, df)

print(df_imputed1)
df_interval_imputed2 <- df_imputed2 %>%
            group_by(interval) %>%
            summarize(avg_steps_per_interval = mean(steps, na.rm = TRUE))

# with(df_interval_imputed2,plot(x= interval, 
#                       y = avg_steps_per_interval, type="l"))

```


## Are there differences in activity patterns between weekdays and weekends?





## Helper functions loaded before analysis
```{r echo=TRUE}
# Helper-functions OBS! will be echoed in the end of the document
# we have gap in the timeseries. Motivation see playground.R
convert_interval_values <- function(interval){
  # 0 5 .... 55 100 105 ... converts to 0 5 .... 55 60 65 ...
  a <- interval %/% 100
  b <- interval %% 100 
  res <- a * 60 + b
  return(res)
}

# Convert back to original value on interval
convert_interval_back2original <- function(interval) {
  # 55 converts to 55
  # 60 converts to 100
  # 125 converts to 205
  a <- interval %/% 60
  b <- interval %% 60 
  res <- a * 100 + b
  return(res)
  
  # lut provides key - value pairs to use to impute "steps" data
imputedata <- function(lut , df) {
  # lut is a dataframe providing suitable replacement for missing values 
  names(lut) <- c("key", "value")
  na_ind <- is.na(df$steps)
  intervals2impute <- df[na_ind, "interval"]
  to_impute <- lut[intervals2impute %in% lut$key ,"value"]
  df$steps[na_ind] <- to_impute
  df
  to_impute
}

  
}

